How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-15 talk/dyla2007/dyla.tex
More reviewer comments.  Check in new pdf.

The observation that limitations of a C-based implementation of a dynamic
language leads to the emergence of additional implementations is clear
in the case of Python.  The reference implementation, \implname{CPython}
\cite{cpy251}, is a simple recursive interpreter.  \implname{Stackless
Python} \cite{stackless} is a fork that adds micro-threading
capabilities to Python. One of the reasons for not incorporating it back
into CPython was that it was felt that this would make the
implementation too complex. Another implementation of the Python
language is \implname{Psyco} \cite{psyco-software}, which adds a
JIT-compiler to CPython.  Finally, \implname{Jython} is a
re-implementation for the Java VM and \implname{IronPython} for
the CLR.  All of these need to be kept synchronized with the relatively fast
evolution of the language.

@@

\item
\emph{Better tools:} Better IDEs, debuggers and profilers.

@@

\noindent
The central theme of the benefits of OO VMs is the ability to
implement certain hard things only once and share the benefits between
all language implementations on top of the OO VM.  At a closer look,
some of these advantages are only partially true in practice:

@@

\item
\emph{Ease of implementation:} This point is disputable. On the one hand, OO
VMs typically allow the language implementer to start at a higher level. On the
other hand, they also enforce a specific object and execution model. This means
that the concepts of the implemented language need to be mapped to the
execution model of the underlying VM, which may or may not be easy, depending very
much on the language in question.

@@

\item
This meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine or integrates it into an existing OO VM.
The necessary support code -- e.g. the garbage collector when
targeting C -- is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

@@

Writing a good garbage collector remains hard, though.  At least, it is
easy to experiment with various kind of GCs, so we started by just using
the conservative Boehm \cite{Boehm} collector for C and moved up to a
range of simple custom collectors -- reference counting, mark-and-sweep,
etc.  Ultimately, though, more advanced GCs will be needed to get the
best performance.  It seems that RPython, enhanced with support for
direct address manipulations, is a good language for writing GCs, so it
would be possible for a GC expert to write one for our translation
framework.  However, this is not the only way to obtain good GCs:
existing GCs can also be reused.  Good candidates are the GCs written in
the Jikes RVM \cite{JikesGC}.  As they are in Java, it should be
relatively straightforward to add a translation step that turns one of
them into RPython (or directly into our RPython-level intermediate
representation) and integrates it with the rest of the program being
translated.

>>>>

The observation that limitations of a C-based implementation of a dynamic
language leads to the emergence of additional implementations is clear
in the case of Python.  The reference implementation, \implname{CPython}
\cite{cpy251}, is a simple recursive interpreter.  \implname{Stackless
Python} \cite{stackless} is a fork that adds micro-threading
capabilities to CPython. One of the reasons for not incorporating it back
into CPython was that it was felt that this would make the
implementation too complex. Another implementation of the Python
language is \implname{Psyco} \cite{psyco-software}, an extension of
CPython which adds a JIT-compiler.  Finally, \implname{Jython} is a
re-implementation for the Java VM and \implname{IronPython} for
the CLR.  All of these ultimately need to be kept synchronized with the
relatively fast evolution of the language.

@@

\item
\emph{Better tools:} Better and cross-language IDEs, debuggers and profilers.

@@

\noindent
The central theme of the benefits of OO VMs is the ability to
implement certain hard things only once and share the benefits between
all language implementations on top of the OO VM.  At a closer look,
some of these advantages are not quite true in practice:

@@

\item
\emph{Ease of implementation:} This point is disputable. On the one hand, OO
VMs typically allow the language implementer to start at a higher level. On the
other hand, they also enforce a specific object and execution model. This means
that the concepts of the implemented language need to be mapped to the
execution model of the underlying VM, which may or may not be easy, depending very
much on the language in question.  In many cases, the mismatch between
the abstractions provided by the underlying VM and the semantics of the
dynamic language is too deep to allow a natural and/or efficient
implementation.

@@

\item
This meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine or integrates it into an existing OO VM.
The necessary support code -- e.g. the garbage collector when
targeting C -- is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; if needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

@@

Writing a good garbage collector remains hard, though.  At least, it is
easy to experiment with various kind of GCs, so we started by just using
the conservative Boehm \cite{Boehm} collector for C and moved up to a
range of simple custom collectors -- reference counting, mark-and-sweep,
etc.  Ultimately, though, more advanced GCs will be needed to get the
best performance.  It seems that RPython, enhanced with support for
direct address manipulations, is a good language for writing GCs, so it
would be possible for a GC expert to write one for our translation
framework.  However, this is not the only way to obtain good GCs:
existing GCs can also be reused.  Good candidates are the GCs written in
the Jikes RVM \cite{JikesGC}.  As they are written in Java, it should be
relatively straightforward to design a translation step that turns them
into RPython (or directly into our RPython-level intermediate
representation).  Our translation toolchain could then integrate any one
of these GCs into the VMs it generates.
